---
title: "Exploratory data analysis"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# Lesson preamble:

> ### Lesson objectives:
>
> -   Implications of (not) understanding your data
>     -   How did you collect your data?
>     -   What are the properties of your data?
> -   Exploring and asking questions about your data with
>     graphing/visualization
> -   Using insights from exploratory analysis to clean up data:
>     -   Dealing with unusual values/outliers
>     -   Dealing with missing values (NAs)
>
> ### Lesson outline:
>
> Total lesson time: 2 hours
>
> -   Data properties, initial predictions (15 min)
> -   Plotting and exploring data (45 min)
> -   Dealing with unusual values (15 min)
> -   Re-connecting with our predictions (30 min)
> -   Dealing with missing values (15 min)

------------------------------------------------------------------------

# Introduction

Exploratory data analysis is your exciting first look at your data! It's
a chance to develop a better understanding of the variables in your data
set and the relationships between them. You can check your assumptions,
find outliers, and possible errors. But THEN you'll get to ask your
questions! Yay!!

You *need* to understand your data you before you analyze it.

1.  What kind of data is it?

2.  What variation is present in my data?

3.  Are there any data points with values beyond the limits I
    anticipated?

4.  Do you notice any patterns?

The patterns you see can lead you to exciting new questions you may not
have anticipated!

# Setup

We'll use what you've learned in past lectures about summarizing and
visualizing data with dplyr and ggplot to get to know some data!

```{r library, message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r download, eval=FALSE}
download.file("https://uoftcoders.github.io/rcourse/data/pseudo.ara.busco", 
              "pseudo.ara.busco")
download.file("https://uoftcoders.github.io/rcourse/data/pseudo.LTRs", 
              "pseudo.LTRs")
download.file("https://uoftcoders.github.io/rcourse/data/pseudoMol_Kdist.txt",
              "pseudoMol_Kdist.txt")
```

We're going to load the genomic data we have on the frequency of genes,
the frequency of a type of repetitive element (LTRs stands for Long
Terminal Repeat - there's some more info on them coming up in the
'predictions' section), and the approximate evolutionary age of those
repetitive elements.

```{r import}
geneDensity <- read_tsv("pseudo.ara.busco", 
                        col_names = c("chromosome", "start", "end", "winNum", 
                                      "numElements", "numBases", "winSize", 
                                      "density"))

ltrDensity <- read_tsv("pseudo.LTRs", 
                       col_names = c("chromosome", "start", "end", "winNum", 
                                     "numElements", "numBases", "winSize", 
                                     "density"))

ltrAge <- read_tsv("pseudoMol_Kdist.txt", col_names=TRUE)
```

We're using "read_tsv" because the columns in this file are separated by
tabs instead of commas or white space. The LTR age data
(pseudoMol_Kdist.txt) already has column names, but the other two data
sets need some more information

# What is my data, actually?

Before we do anything else, we have to think about where this came from
& whether the data is appropriate for the kinds of questions we might
have.

This data describes a two of the genetic units (we'll call them
"elements") that live in one plant genome: a set of highly conserved
genes and one type of transposon (a "selfish" gene that makes copies of
itself at the expense of its host genome). The chromosomes have been
broken down into 1Mb pieces ("windows") that overlap each other. In each
window, we know the number and size (base pairs occupied) of the
conserved genes and transposons.

## Predictions

It's always good to lay out your hypotheses first. It can help you
figure out how you need to assemble your data in order to test those
predictions effectively.

1.  In areas where gene density is high, LTR density is low
    -   LTRs are a type of **transposable element**, aka "genomic
        parasite"
        -   They make copies of themselves at the expense of their host
            genome
        -   They make up a large portion of plant genomes (can be
            \>40%!)
        -   The host genome wants to prevent them from replicating
    -   Certain regions of a chromosome are more tightly wound up with
        histones
        -   This makes them less accessible to molecular machinery
        -   If polymerases aren't likely to access a region, the region
            can't be expressed
        -   If a region is unexpressed, you don't want genes there!!
        -   LTRs tend to accumulate in these regions
    -   More accessible, active regions of a chromosome have higher gene
        content
        -   These regions can be expressed effectively!
        -   LTRs that insert into these regions have a worse impact on
            the host
    -   Other factors like recombination rate and methylation also
        support this pattern
2.  In areas where gene density is high, LTR age will be high (old, not
    transposing anymore)
    -   There won't be many new deleterious LTR insertions
        -   Few young LTRs
    -   The LTRs that are present in those regions can't have lethal
        effects
        -   If they're there, their effects are unlikely to have
            terrible effects on fitness
        -   Some transposable elements have been "domesticated" by their
            hosts
        -   This all contributes to the likelihood that LTRs present
            can/have persisted
        -   LTRs present are more likely to be older
3.  The sex chromosome (LG_X) will have higher LTR density
    -   Larger proportions of sex chromosomes are less accessible
    -   Sex chromosomes experience lower rates of recombination relative
        to autosomes
        -   Also correlated with higher transposon density and lower
            gene density
    -   These trends are more true for non-recombining Y chromosomes
        than X chromosomes
        -   Recombination can occur between the two X chromosomes in
            females

## First Peek

First, let's just take a quick look at the gene density data set and ask
ourselves what we're dealing with. On a very basic level, what kind of
variables do we have?

What is one way to view a data frame?

```{r message=FALSE, warning=FALSE}
#head(geneDensity) # prints the first 6 rows
#tail(geneDensity) #prints the last 6 rows
glimpse(geneDensity) #prints number of rows and columns, column names, types, and several entries
```

What are your first impressions of the data?

Which variables will be relevant for testing our predictions?

**Something to note here: in this data set, there are a number of things
that will be clearly**

## Basic Variable Categories

Common variable types:

-   Independent vs dependent
-   Continuous vs discrete
-   Qualitative: categorical/nominal, ranked/ordinal, dichotomous
-   Quantitative: interval, ratio

This matters because the type of data tells us the appropriate way to
visualize it:

-   Qualitative data: pie charts or bar charts
-   Quantitative data: histograms, box plots

# Visualizing Your Data

## Describing Patterns in Histograms

For a given variable, you're generally looking at the range of values
and where the majority of the data lies. This gives you an idea of the
**distribution** of your data. As you're aware, many statistical tests
make assumptions about the distribution of your input data - it's very
important to make note of the shapely properties of your data.

-   Average (mean, median, mode)
-   Range (max, min)
-   Skewness: how symmetrical is your data around the average?
    -   Classic bell curve has a skew of zero
    -   If your data isn't symmetrical, that can give you important
        info!
    -   Skewed distributions aren't likely to be normally distributed
-   Kurtosis: how sharp is your central peak?
    -   If your distribution is basically flat, its kurtosis is negative
    -   If your distribution has a huge spike, its kurtosis will be
        positive

## Qualitative Data with Histograms

Histograms are great for qualitative data because they visualize the
number of times a given value appears in your data.

## Quantitative Data with Histograms

Histograms provide an important view into continuous data, providing
that you tell ggplot how to group your data into discrete bins. Here, we
can look at our data's values for gene density. This density is a
measurement of the number of base pairs in a 1Mb window that are part of
a gene divided by the total number of base pairs in the window (1 000
000).

```{r}
ggplot(geneDensity, aes(density)) +
  geom_histogram(binwidth = 0.01) + 
  labs(title = "Distribution of gene density values",
       x = "Gene density", y = "Count (bin size = 0.01)") # Adding labels helps!
```

What are some words you'd use to describe this distribution?

Does this change

## Binning Quantitative Data

When you're subsetting continuous data into discrete bin widths, it's
important to try out different values because different bin sizes can
give *vastly* different impressions of your data's distribution.

```{r}
ggplot(geneDensity, aes(density)) +
  geom_histogram(binwidth = 0.001) +  # Teeny tiny bins
  labs(title = "Distribution of gene density values",
       x = "Gene density", y = "Count (bin size = 0.001)")

ggplot(geneDensity, aes(density)) +
  geom_histogram(binwidth = 0.1) +  # Huge bins! (for this data)
  labs(title = "Distribution of gene density values",
       x = "Gene density", y = "Count (bin size = 0.1)")
```

It's also interesting to see whether your data's distribution is
different for any of the categories you're looking at. Is there greater
variation in height among women vs humans as a whole? (Do be careful
with this, because looking for patterns by poking your data into a bunch
of different subsets will basically guarantee you'll find *a* pattern,
whether or not it's biologically relevant.)

### Histogram for One Chromosome

Let's see whether the gene density on one of the autosomes (how about
LG_2) fits the general pattern.

(Based on our initial hypotheses, would you predict that it would?)

It is important to consider how your predictions may affect the way you
filter your data, so be mindful about tweaking parameters (like bin
width) to fit the output you expect!

```{r}
geneDensity %>%
  filter(chromosome == "LG_2") %>%
  ggplot(aes(density)) +
  geom_histogram(binwidth = 0.01) +
  labs(title = "Distribution of gene density values on LG_2",
       x = "Gene density", y = "Count")
```

The range for the x axis is much smaller! The maximum gene density here
(\~12%) is much smaller than the highest value in the full genome data
set (\~40/50%).

(Why might this be?)

One of the aspects of your data that you can't visualize well with a
histogram is whether there are any values that exceed the limits you
expected for your data.

## Scatterplots & Box plots

### More info! Less bias!

With quantitative data, we can get more information by looking at
scatterplots and box plots. Not only are they immune to bin size bias,
they can help us find outliers and let us make initial visual
comparisons of averages across categories.

### Visualize raw data as a scatterplot

We know that "chromosome" is a categorical, independent variable
appropriate for our X axis and that "density" is a continuous, dependent
variable that will be appropriate for the Y.

```{r}
ggplot(geneDensity, aes(x = chromosome, y = density)) +
  geom_point() +
  labs(title = "Comparison of gene density across chromosomes",
       x = "Chromosome", y = "Gene density")
```

Already, we can see that there different maximum gene density values on
each chromosome. Because the points are overlapping, it's hard to
evaluate what the average or skewness might be for any of the
categories.

### Boxplots for better comparisons

Because boxplots display the median and quartile limits, it's much
easier to evaluate the properties of the distribution.

```{r}
ggplot(geneDensity, aes(x = chromosome, y = density)) +
  geom_boxplot() +
  labs(title = "Comparison of gene density across chromosomes",
       x = "Chromosome", y = "Gene density")
```

There's definitely a value that jumps out immediately. It's stretching
the scale of the Y axis so that it's hard to effectively compare the
medians of each of the chromosomes.

Before we officially decide what to do with this outlier, we'll visually
set it aside for now by re-scaling our Y axis, which we've already
learned how to do!

```{r}
ggplot(geneDensity, aes(x = chromosome, y = density)) +
  geom_boxplot() +
  ylim(0, 0.125) + #other methods possible
  labs(title = "Comparison of gene density across chromosomes",
       x = "Chromosome", y = "Gene density")
```

Look at that handy warning! It lets us know that one value was thrown
out: "removed 1 rows". This view helps us to get a better general
understanding of how this categorical "chromosome" value might relate to
gene density. However! It's important not to throw out data unless you
have good reason to believe it doesn't belong in the data set.

Bonus note: you can use the coord_cartesian function instead of ylim. It
won't warn you if any of your data points are beyond the limits of the
axes, though.

### What about the other variables?

> This data describes a few of the genetic "bits" (we generally call
> them "elements") live in one plant genome. The chromosomes have been
> broken down into 1Mb pieces ("windows") that overlap each other and
> the contents of each window have been averaged. We've got information
> on the density of conserved genes and one type of transposon for each
> window. Additionally, we have the evolutionary age for those
> transposons.

Average number of genes in bins along chromosomes.

Definitely more interesting to compare across the categories built into
our data (here, chromosomes) to see how the gene density looks in each
one separately. We can see whether the global pattern is present in each
category. But how can we get all that info in one graph??

First step is to ask ourselves what we currently have in our data. If
our category for comparison is chromosome, what independent variables
are shared among them that could facilitate comparison of the dependent
gene density variable?

```{r}
head(geneDensity)
```

Start, end, and winNum would all be reasonable proxies for position
along the chromosome.

```{r}
geneDensity %>%
  filter(chromosome == "LG_2") %>%
  ggplot(aes(x = start, y = density)) +
  geom_point() +
  labs(title = "Comparison of gene density along LG_2",
       x = "Chromosomal position (bp)", y = "Gene density")
```

This gives us an overview of how many of the conserved genes are found
in which region of this LG_2 chromosome.

To be able to compare all the chromosomes at the same time, we can split
our graph into "facets" so there's one per chromosome, as you've learned
how to in the last lecture.

```{r}
ggplot(geneDensity, aes(x=start, y=density)) +
  geom_point() +
  labs(title="Comparison of gene density across chromosomes",
       x="Chromosomal position (bp)", y="Gene density") +
  facet_wrap( vars(chromosome) )
```

Because not all of the chromosomes are the same length, the data appears
more squished in some of the panels. We can adjust that by telling facet
wrap to scale the X axis per-panel instead of globally.

If we want to be able to visually compare the densities across
chromosomes, we should *not* allow the Y axis to scale freely. We can,
however, set a limit for the Y axis values, as we've done before.

Use different command for scaling the Y axis

```{r}
ggplot(geneDensity, aes(x = start, y = density)) +
  geom_point() +
  coord_cartesian( ylim = c(0,0.13) ) +
  labs(title = "Comparison of gene density across chromosomes",
       x = "Chromosomal position (bp)", y = "Gene density") +
  facet_wrap( vars(chromosome), scales = "free_x" )
```

Cool, eh?? The chromosomes have very different patterns! The range and
distribution of values differs considerably!

What are some reasons for gene density to change along a chromosome?

-   Centromeres are mostly made up of repeats - very low gene content
    -   Centromeres can be in the middle or near one end of a chromosome

    -   Where do you think the centromeres are in these chromosomes?
-   Certain regions of a chromosome are more tightly wound up with
    histones
    -   Makes them less accessible to molecular machinery

    -   If polymerases don't reach a region, that region can't be
        expressed

    -   If a region is unexpressed, you don't want genes there!

    -   Centromeres are generally one of these 'inactive' regions
-   More accessible, active regions of a chromosome have higher gene
    content
    -   These regions are generally along chromosome arms

### Boxplot augmentations

There are a few additional things we can do that might make boxplots
even more informative for our data.

-   **Violin** plots - boxplots but with curves instead of boxes
-   Adding a **scatterplot** behind the boxplot
-   Adding "**jitter**" to scatterplots so the points are offset
    -   Additionally, you can make the points more transparent (change
        the **alpha** value)
-   You can also add a **trend line** to help you visualize potential
    relationships

```{r}
ggplot(geneDensity, aes(x = chromosome, y = density)) +
  geom_point(alpha = 0.1, position = "jitter") +
  geom_violin() + 
  labs(title = "Comparison of gene density across chromosomes",
       x = "Chromosome", y = "Gene density")
```

Making the points more transparent gives us a better idea of what
density values are most common. You can see this at the bottom of the
graph, where the points don't look transparent at all - *so* many data
points!!

## Challenge!

How could you visualize the LTR data across chromosomes? Don't forget to
use axis labels.

What is the range of LTR density for the LG_2 chromosome?

```{r}
ggplot(ltrDensity, aes(x = chromosome, y = density)) +
  geom_violin() + #boxplot also valid
  geom_point(alpha = 0.01, position = "jitter") +
  labs(title = "Comparison of LTR density across Chromosomes",
       x = "Chromosome", y = "LTR density")

ltrDensity %>%
  group_by(chromosome) %>%
  summarize(mean = mean(density), median = median(density), 
            n = n(), max = max(density))
```

Now it's time to start thinking about what to do with rebellious
outliers!

# Outliers

## But why are you like this?

There could be many reasons why your data has values that exceed the
limits you expected it would have. It basically comes down to **error**,
whether in your data or in the expectation you had for its limits.
Consider error *carefully*.

-   Incorrect prediction of what the limits should be
    -   Maybe your study system has different qualities than literature
        spp.
-   **Systematic error** is predictable and affects a measurement's
    accuracy
    -   Incorrectly calibrated lab equipment (pH meter, thermometer,
        etc.)
    -   Genomics - your gene annotation can be biased by repetitive
        elements
    -   Can be very difficult to compensate for this kind of error
-   **Random error** affects measurement precision (think significant
    figures)
    -   Writing down measurements incorrectly in your notes
    -   A lab scale can only weigh samples to a given decimal point
    -   Simple human fallibility when it comes to assessing measurements
    -   Take multiple measurements and average the results to compensate
-   Common sources
    -   This (exploratory data analysis) is a great time to look for
        issues!
    -   Error in previous analysis steps (code that produced LTR age
        estimates)
    -   Erroneous assumptions about your sample sites or methods
-   Don't just throw out a point because you *can* ascribe it to error

IMPORTANT NOTE: if you do end up removing *any* data, you **MUST
disclose the removal** and your justification for the removal. Your
reasons could be great, but you need to make sure your audience has
access to those reasons.

This consideration of weird values brings up an interesting point:
you're doing these checks because your values are *different* than what
you expected. It's important to think about analytical 'controls' to
look for potential errors even when your data looks the way you *expect*
it to! Steps towards this can be as simple as sharing your code
publicly.

## Let's take a look!

We had that weird really high gene density value on the LG_X chromosome.
Let's look at what's happening there.

```{r highdensity}
filter(geneDensity, density > 0.13)
```

What do the other variables tell us about this data point?

This data point has a really high winNum, so it's likely located near
the end of the chromosome. But importantly, our windows are supposed to
be 1Mb in size (1 000 000 value in the winSize column). The winSize
value for this outlier is tiny in comparison!!

## Wholesome thoughts about your data

Averages -- how was your data collected & what biases might be inherent?
The data I'm showing you is a pretty clear example of how important (and
difficult) it is to understand what the variables mean in your data
sets. What might cause outliers in the kind of data you're interested
in?

That last look showed us that it's definitely very important to consider
our data as a whole: to think not only about the variables relevant to
our hypotheses, but the way in which it was collected (how that was
reflected in the other properties of the data).

So. Let's try to understand more about the interaction between gene
density and window size in the rest of our data. Visualization time!

```{r}
ggplot(geneDensity, aes(x = start,y = winSize, colour = chromosome)) +
  geom_point() +
  labs(title = "Window sizes along the chromosome",
       x = "Chromosomal position (bp)", y = "Window size (bp)")
```

It looks like *all* of the chromosomes get this trail-off in window size
near their ends. This is not what we expected!! All of the squish at the
end is basically just error from a previous analysis.

## Challenge!

Create a category based on window size: density as either belonging in a
"small window" or a "normal window". We can create a new "winCategory"
variable using mutate() and assign the value of "small" to windows with
winSize less than 1Mb and "normal" to all the other windows (which we
expect will have a value of 1Mb).

```{r}
geneDensity2 <- geneDensity %>%
  mutate( winCategory = case_when(winSize<1000000 ~ "small",
                                  TRUE ~ "normal") ) %>%
  group_by(winCategory, chromosome)

summarize(geneDensity2,
          mean = mean(density), median = median(density), n = n(), 
          max = max(density), sd = sd(density), .groups = "keep")
```

What can we take away from this table?

The n values are considerably larger for the normal-sized windows group.
LG_4 and LG_N had 0 gene density in their small windows but have some of
the highest median gene densities in the normal-sized windows.

The standard deviation of the small windows is much higher. Is that what
we would expect for that data category? Perhaps most importantly for our
purposes, the **mean** and **median** are quite different. These smaller
windows have considerably different values.

What does this look like in an actual plot? This is going to take a bit
of black magic in the form of two separate calls to geom_boxplot(). The
first will use all the windows (setting it to colour values by 'all')
and the second will actually create (and colour) different box plots
based on their winCategory value.

```{r}
ggplot(geneDensity2, aes(x = chromosome, y = density, colour = winCategory)) +
  geom_boxplot( aes(x = chromosome, y = density, colour = "all") ) + 
  geom_boxplot() +
  ylim(0, 0.125) +
  labs(title="Visualizing gene density across window size and chromosome",
       x="Chromosome", y="Gene density", colour="Window\nCategory")
```

The small window values seem quite different than the global gene
density results!

Why do you think this might be? Looking back on the summaries, we can
see that there aren't many data points in the 'small' window category.

In conclusion!!

These small windows do seem to contain interesting information, but they
are over-weighted given the amount of information they're based on.
Based on the analysis conducted to create the windows, it *might* be
appropriate to discard the small windows on the ends of the chromosomes.
Each windowed data point is weighted equally, even though these smaller
windows contain less information, which creates a bias.

What do **you** think is the most appropriate way to deal with this
data?

Is there a way to weight the gene density by window size?

# So what does this mean for our predictions?

Right. The reason we collected this data in the first place!

1.  In areas where gene density is high, LTR density is low
2.  In areas where gene density is high, LTR age will be high
3.  The sex chromosome (LG_X) will have higher LTR density

Note: *preparing* data for analysis is generally the most time-consuming
part of any project. Establishing what you need to do with your data in
order to test your hypotheses & thoroughly exploring your data and its
properties are *extremely* useful steps in this process.

## How do we explore these questions?

We need to relate gene density, LTR density, and LTR age. Do we have
this data?

Is it currently in a form where we can make comparisons?

Based on the properties of our gene and LTR density data sets, what are
the shared "units"? Essentially, what are we trying to compare within
and among each chromosome?

```{r}
head(geneDensity)

as_tibble(ltrDensity)
```

The basic "unit" in this data is the 1Mb window. Because this is shared
across the two data sets, we can use it to join them together.
Excellent!

What about the LTR age data set?

```{r}
as_tibble(ltrAge)
```

This data was prepared differently, so it doesn't have the same 'window'
units. It does contain chromosomal position information, however, which
we can use to make some preliminary comparisons.

### Actual wrangling

We're also going to pull out only the variables we now know we'll need
(what's shared among the data sets and what will be used to try and test
our predictions), just because of how large this data frame will be.
It's *not* a good idea to do this before looking at all the variables
together.

```{r}
simpleGeneDensity <- geneDensity %>%
  mutate(elementType = "gene") %>%
  select(chromosome, start, elementType, density)

simpleLTRdensity <- ltrDensity %>%
  mutate(elementType = "LTR") %>%
  select(chromosome, start, elementType, density)

head(simpleLTRdensity)
```

At this point, are these data "long" or "wide"? #throwback

### ? Knowledge Check Challenge

Join the two data sets (simpleLTRdensity and simpleGeneDensity) into one
data frame called "densities". As a bonus, try mutating the *start*
variable so that it's measured in 10kb increments instead of 1bp. This
will just make our X axis labels are easier to interpret.

```{r}
densities <- full_join(simpleLTRdensity, simpleGeneDensity,
                       by = c("chromosome", "start", "elementType", "density")) %>%
  mutate(start = start / 10000) %>%
  group_by(chromosome, elementType)
  
head(densities)

rm(simpleGeneDensity, simpleLTRdensity)
```

We've got two independent categorical variables, an independent
numerical variable, and a dependent numerical variable. It's beautiful.

## Is gene density high when LTR density is low? (hyp #1)

What variables do we want to plot?

-   Chromosome
-   Start position (bp)
-   Element type
-   Element density
-   

### Challenge

Of the plot types we've used so far, what would you use to try and
compare gene densities along the chromosomal positions on each
chromosome?

```{r histogram}
ggplot(densities, aes(x = density, fill = elementType)) +
  geom_histogram( binwidth = 0.03 ) +
  facet_wrap( vars(chromosome), scales = "free_y" ) +
  coord_cartesian(xlim = c(0,0.6)) +
  labs(x = "Element Density", y = "Count", fill = "Element\nType",
       title = "Element densities among chromosomes")
```

Poking at the histogram shows us some interesting things about
differences in the frequencies of LTRs and genes. Gene values have
extremely high kurtosis near 0. LG_4 may have the highest median/mode
LTR density.

The Y axis can be free-scaled here because all of the counts are based
on the size of their chromosome. We don't want one chromosome to seem
like it has a much higher LTR count just because it has more windows
(greater n) than the other chromosomes.

This was an interesting plot, but it **compares densities across
chromosomes** more than it looks at **differences in LTR/gene patterns
within chromosomes**.

```{r scatterplot}
ggplot(densities, aes(x = start,y = density,colour = elementType)) +
  geom_point(alpha = 0.3) +
  facet_wrap( vars(chromosome), scales = "free_x" ) +
  ylim(0, 0.5) +
  labs(title = "Element densities along chromosomes",
       x = "Chromosomal position (10kb)", y = "Element density",
       colour = "Element\nType")
```

This looks like the kind of information we want! If we squint, we can
almost see that increases in gene density seem to correlate with
decreases in LTR density.

If you can remember how to add a *smooth* line to show broad patterns,
this will be the easiest view.

```{r trendline-advanced}
ggplot(densities, aes(x = start,y = density,colour = elementType)) +
  geom_smooth() +
  ylim(0, 0.4) +
  facet_wrap( vars(chromosome), scales = "free_x" ) +
  labs(title = "Element densities along chromosomes",
       x = "Chromosomal position (10kb)", y = "Element density",
       colour = "Element\nType")
```

Broadly, we can see that when the LTR density plummets, gene density
smudges upward.

## Is gene density high when LTR age is high? (hyp #2)

Let's take a look at the LTR age data.

```{r}
head(ltrAge)
```

Some of the variable names are different (K2P is our measure of age) but
are really familiar to the other data we've been analyzing (chrom
instead of chromosome). Let's see if the LTR age data looks anything
like our gene and LTR density data.

```{r}
ggplot(ltrAge, aes(x = start,y = K2P)) +
  geom_point(alpha = 0.1) +
  facet_wrap( vars(chrom), scales = "free_x" ) +
  labs(title = "LTR age along chromosomes",
       x = "Chromosomal position", y = "LTR age (K2P)")
```

These clouds of points are really hard to understand. We can try using
geom_smooth to see if it can reveal what's going on in these clouds. The
one argument we'll make note of right now is *n*, which tells
geom_smooth how many points along the x it should be using to make its
average. Because of how big our data is, we'll give it a smaller value
so it doesn't take forever to plot.

```{r}
ggplot(ltrAge, aes(x = start,y = K2P)) +
  geom_point(alpha = 0.1) +
  geom_smooth(n = 50) + # try different values (think histogram bin widths)
  facet_wrap( vars(chrom), scales = "free_x" ) +
  labs(title = "LTR age along chromosomes",
       x = "Chromosomal position", y = "LTR age (K2P)")
```

Well, what can we take away from this visualization? Not much. It seems
pretty clear that the average LTR is old (darkness at the bottom of the
clouds). The few younger LTRs near the top of the plots might have
useful information for us, given that LTR "reproduction" (transposition)
is generally rare.

## Discretizing our ages

We can try binning LTRs based on their age, to see if the youngest LTRs
are able to be "born" in gene-dense regions. Note: don't categorize
numerical data like this without considering what information you're
losing!

```{r}
ggplot(ltrAge, aes( x = start / 10000,
                    y = K2P,
                    colour = cut(K2P, 4,
                               labels = c("youngest", "young",
                                          "old", "oldest")) 
                    )) +
  geom_point(alpha = 0.5) +
  #geom_smooth() + #just aren't enough points in the younger categories
  facet_wrap( vars(chrom), scales = "free_x" ) +
  labs(title = "LTR age along chromosomes",
       x = "Chromosomal position (10Mb)", y = "LTR age (K2P)",
       colour = "Age Class")
```

There are few young LTRs, but their positions match with the pattern of
our overall TE density plot. The positions of the young LTRs aren't
close to gene-dense regions on the autosomes, though interestingly there
are a few on the right-hand side of LG_X, which was the most gene-dense
region of our sex chormosome!

That was cool. Definitely **not** a statistical test of correlations
between LTR age and gene density, but the fact that sub-sections of this
data behave quite differently is really interesting to see!

## Missing values in gene density comparison

You might remember that in last lecture, when you encountered missing
values in your data, you could replace them with 0. If we want to put
our LTR age data into the same kind of windows that the gene density
data is in, we're going to have *some windows without any age data*. Do
you think that setting LTR age to 0 would be a good way to handle the
windows with missing age data?

```{r}
windowedAges <- ltrAge %>%
  mutate(chromosome = chrom,            # get the "chromosome" col name to match
         age = K2P,                     # might as well give a better name
         winNum = floor(start/20000)) %>% # bin the start (20Mb) & round 
  select(chromosome, winNum, age)
  
genesPlusAges <- geneDensity %>%
  select("chromosome", "winNum", "density") %>%
  full_join(windowedAges, by = c("chromosome", "winNum") )

as_tibble(genesPlusAges)
```

Just for the record, this is not a great way to window data.

But! Look what happens when you change the join method! Got more missing
data... Maybe it'll be easier to understand if we plot it.

```{r}
findMissing <- genesPlusAges %>%
  mutate(noDense = is.na(density),
         noAges = is.na(age))

ggplot(findMissing) +
  geom_point( aes(x = winNum, y = density, colour = noAges) ) +
  coord_cartesian( ylim = c(0,0.15) ) +
  facet_wrap( vars(chromosome), scales = "free_x" ) +
  labs(title = "Missing gene density data",
       x = "Window number", y = "Gene density", colour = "Missing ages!")

ggplot(findMissing) +
  geom_point( aes(x = winNum, y = age, colour = noDense) ) +
  facet_wrap( vars(chromosome), scales = "free_x" ) +
  labs(title = "Missing LTR age data",
       x = "Window number", y = "LTR age", colour = "Missing densitiy!")
```

Ooph. We're only missing gene density data for 3 windows, but we're
missing out on LTR age data for around 15k windows... Is there anything
we can do about it?

Assigning a value of 0 to our LTR ages is asserting we have different
data than we actually have: that we know there's an LTR in that position
and that it's an incredibly recent insertion. But what if we replace the
missing age points with the mean LTR age?

```{r}
findMissing <- genesPlusAges %>%
  mutate(noDense = is.na(density),
         noAges = is.na(age),
         rplAge = replace_na( age,mean(age,na.rm=TRUE) ),
         rplDensity = replace_na( density,mean(density,na.rm=TRUE) ))

sum( is.na(findMissing$rplDensity) )
sum( is.na(findMissing$rplAge) )

ggplot(findMissing) +
  geom_point(alpha = 0.5,
             aes(x = winNum, y = rplAge,
                 colour = cut(rplAge, 3, labels = c("youngest", "middle", "oldest")))) +
  geom_col(aes(x = winNum, y = rplDensity)) +
  ylim(c(0,0.2)) +
  facet_wrap(vars(chromosome), scales = "free_x" ) +
  labs(title = "Gene density and LTR age along chromosomes",
       x = "Chromosomal position (window number)", y = "Altered ages",
       colour = "Age Class")
```

No more warnings about missing data! Woo! We can feel whole again! On
the other hand, replacing these values didn't actually help us improve
our understanding of our age-density prediction...

We should actually plot a window's LTR age against its gene density!
First, let's look at the general pattern with the adjusted ages (when we
replaced NA with the mean value for the variable).

```{r}
ggplot(findMissing) +
  geom_point(alpha = 0.5,
             aes(x = rplDensity, y = rplAge,
                 colour = cut(rplAge, 3, labels = c("youngest", "middle", "oldest")))) +
  coord_cartesian( xlim = c(0, 0.13) ) +    # being mindful of our outlier
  facet_wrap( vars(chromosome), scales="free_x" ) +
  labs(title="Gene density vs LTR age post-adjustment",
       x="Gene density (adjusted)", y="LTR age (adjusted)",
       colour="Age Class")
```

That's pretty dang cool! We do see a general downward trend - as gene
density increases (going right along the X axis), LTR age tends to
decrease! We can even see some interesting differences among the
chromosomes.

LG_7 is our biggest chromosome and it seems to have a lot of the young
(blue) LTRs. Why do you think that might be? *Generally*, as chromosome
length increases, recombination rate decreases. Recombination rate is
one of the properties we know correlates with the accessibility of
genomic regions. So, we might predict that longer chromosomes have a
larger proportion of 'inactive' regions and a greater number of TEs.

Moving on!

Let's judge our past selves! Let's see the effects of our decision to
replace the missing age data with an average. Was it foolish or as wise
as the time we joined a dragon boat team?

```{r}
ggplot(findMissing) +
  # This time, we'll colour the points based on whether the age should be NA
  geom_point(alpha=0.5, aes(x=rplDensity,y=rplAge, colour=noAges)) +
  coord_cartesian( xlim=c(0,0.13) ) +    # being mindful of our outlier
  facet_wrap( vars(chromosome), scales="free_x" ) +
  labs(title="Gene density vs LTR age post-adjustment",
       x="Gene density (adjusted)", y="LTR age (adjusted)",
       colour="Agelessss")
```

Not too bad! Not mechanistically motivated, but it does show that
replacing missing data with the value we would expect (basically the
definition of the mean) can have fewer consequences than replacing it
with 0 in some contexts.

## Does the sex chromosome (LG_X) have higher LTR density? (hyp #3)

We want to compare LTR densities across chromosomes. How would you do
this?

### Challenge!

Box plot is a good base line. Ask for types of improvements that could
be made to a basic box plot.

```{r}
ggplot(ltrDensity, aes(x = chromosome, y = density)) +
  geom_boxplot() +
  labs(title = "LTR densities among chromosomes",
       x = "Chromosome", y = "LTR density")
```

What unusual things do you notice about this plot?

Does this initial exploration lead us to think that the sex chromosome,
with its reduced rate of recombination, has accumulated more LTRs?

Based on what we know about the gene density data, what would you
suggest we might need to do with this LTR density data?

There is one extreme outlier again - let's see what happens if we colour
the data based on window size.

```{r}
ltrDensity %>%
  ggplot( aes(x = chromosome, y = density, colour = winSize==1000000) ) +
  geom_boxplot() +
  ylim(c(0,0.6)) +
  labs(title = "LTR densities among chromosomes",
       x = "Chromosome", y = "LTR density", colour = "Normal\nWindow\nSize")
```

There are fewer outliers, but the pattern remains the same. GOOD SIGN!
:)
